---
title: "Stat 3333 Final Project"
author: "Elizabeth Flores, Yohari Hernandez, Regis Ouedraogo, Richard Sauve"
date: '2022-12-12'
output:
  word_document: default
  pdf_document: default
subtitle: An Analysis of Excess Death in Early Covid-19 Pandemic
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Data Preparation and Loading

```{r}
# Libraries
library(tidyverse)
library(ggplot2)
```

```{r}
# Importing data
set.seed(2022)
deaths <- read.csv('Excess_Deaths_Associated_with_COVID-19.csv', header=TRUE)
colnames(deaths)[1] <- 'Week.Ending.Date'
deaths$Week.Ending.Date <- as.Date(deaths$Week.Ending.Date)
deaths <- deaths %>% na.omit()

# Data is total excess deaths per week from 2017-01-01-2020-06-30
deaths %>% group_by(Week.Ending.Date) %>% summarise(total = round(mean(Observed.Number))) -> total_deaths

# Filter data by year for Jan-June period
year2018 <- total_deaths %>% filter(Week.Ending.Date <= '2018-06-30' & Week.Ending.Date >= '2018-01-01') %>% mutate(Year = 2018)
year2019 <- total_deaths %>% filter(Week.Ending.Date <= '2019-06-30' & Week.Ending.Date >= '2019-01-01') %>% mutate(Year = 2019)
year2020 <- total_deaths %>% filter(Week.Ending.Date >= '2020-01-01' & Week.Ending.Date <= '2020-06-30') %>% mutate(Year = 2020)

# Combine data for ANOVA
anova <- bind_rows(year2018, year2019, year2020)

# Summary statistics
cat("Sample sizes:\n")
cat("2018:", nrow(year2018), "observations\n")
cat("2019:", nrow(year2019), "observations\n")
cat("2020:", nrow(year2020), "observations\n\n")

cat("Descriptive Statistics:\n")
cat("2018 - Mean:", round(mean(year2018$total), 2), "SD:", round(sd(year2018$total), 2), "\n")
cat("2019 - Mean:", round(mean(year2019$total), 2), "SD:", round(sd(year2019$total), 2), "\n")
cat("2020 - Mean:", round(mean(year2020$total), 2), "SD:", round(sd(year2020$total), 2), "\n")
```

## 2. Exploratory Data Analysis

### Time Series Visualization

```{r, fig.width=20, fig.height=10}
# Looking into data - Time series plots
par(mfrow = c(2, 2))
plot(y = year2018$total, xlab = 'Date', ylab = 'Excess Deaths',x = year2018$Week.Ending.Date, type = 'l', ylim = c(1600, 2700), main = 'Excess deaths 2018 Jan-Jul', lwd = 5, col = 'blue')
plot(y = year2019$total, xlab = 'Date', ylab = 'Excess Deaths',x = year2019$Week.Ending.Date, type = 'l', ylim = c(1600, 2700), main = 'Excess deaths 2019 Jan-Jul', lwd = 5, col = 'green')
plot(y = year2020$total, xlab = 'Date', ylab = 'Excess Deaths',x = year2020$Week.Ending.Date, type = 'l', ylim = c(1600, 2700), main = 'Excess deaths 2020 Jan-Jul (COVID-19)', lwd = 5, col = 'red')

# Combined comparison
plot(y = year2018$total, xlab = 'Date', ylab = 'Excess Deaths', x = year2018$Week.Ending.Date, type = 'l', 
     ylim = c(1600, 2700), main = 'Comparison of Excess Deaths Across Years', lwd = 3, col = 'blue')
lines(y = year2019$total, x = year2019$Week.Ending.Date, lwd = 3, col = 'green')
lines(y = year2020$total, x = year2020$Week.Ending.Date, lwd = 3, col = 'red')
legend('topright', legend = c('2018', '2019', '2020'), col = c('blue', 'green', 'red'), lwd = 3)
```

**Observation:** The 2020 data shows elevated excess deaths compared to 2018 and 2019, particularly starting in March, coinciding with COVID-19 pandemic onset in the US.

### Distribution and Normality Assessment

```{r, fig.width=15, fig.height=10}
par(mfrow = c(3, 2))
hist(year2018$total, main = "2018 Excess Deaths", xlab="Excess Deaths", ylab="Frequency", xlim = c(1800, 2800), breaks = 8, col = 'lightblue')
qqnorm(year2018$total, main = "Q-Q Plot of 2018")
qqline(year2018$total, col = 'red', lwd = 2)

hist(year2019$total, main = "2019 Excess Deaths", xlab="Excess Deaths", ylab="Frequency", xlim = c(1800, 2800), breaks = 5, col = 'lightgreen')
qqnorm(year2019$total, main = "Q-Q Plot of 2019")
qqline(year2019$total, col = 'red', lwd = 2)

hist(year2020$total, main = "2020 Excess Deaths", xlab="Excess Deaths", ylab="Frequency", xlim = c(1800, 2800), breaks = 10, col = 'lightcoral')
qqnorm(year2020$total, main = "Q-Q Plot of 2020")
qqline(year2020$total, col = 'red', lwd = 2)
```

### Formal Normality Testing

```{r}
cat("Shapiro-Wilk Normality Test:\n")
cat("2018: p-value =", round(shapiro.test(year2018$total)$p.value, 4), "\n")
cat("2019: p-value =", round(shapiro.test(year2019$total)$p.value, 4), "\n")
cat("2020: p-value =", round(shapiro.test(year2020$total)$p.value, 4), "\n")
cat("Note: p > 0.05 suggests data are approximately normally distributed\n")
```

## 3. Bootstrap Confidence Intervals

```{r, fig.width=30, fig.height=12}
set.seed(2022)
n2018 <- length(year2018$total)

N <- 10^4

# Vectorized bootstrap sampling (more efficient than loops)
samp2018 <- replicate(N, mean(sample(year2018$total, n2018, replace = TRUE)))
samp2019 <- replicate(N, mean(sample(year2019$total, n2018, replace = TRUE)))
samp2020 <- replicate(N, mean(sample(year2020$total, n2018, replace = TRUE)))

par(mfrow = c(1, 3))
hist(samp2018, main = "2018 Bootstrap Distribution of Mean", xlab="Bootstrap Mean 2018", col = 'lightblue', breaks = 30) 
abline(v = mean(year2018$total), col = "darkblue", lty = 2, lwd = 2)

hist(samp2019, main = "2019 Bootstrap Distribution of Mean", xlab="Bootstrap Mean 2019", col = 'lightgreen', breaks = 30) 
abline(v = mean(year2019$total), col = "darkgreen", lty = 2, lwd = 2)

hist(samp2020, main = "2020 Bootstrap Distribution of Mean", xlab="Bootstrap Mean 2020", col = 'lightcoral', breaks = 30) 
abline(v = mean(year2020$total), col = "darkred", lty = 2, lwd = 2)
```

### 95% Confidence Intervals for Mean Excess Deaths

```{r}
cat("95% Confidence Intervals (Bootstrap):\n\n")

ci2018 <- quantile(samp2018, c(0.025, 0.975))
cat("2018: [", round(ci2018[1], 2), ",", round(ci2018[2], 2), "]\n")

ci2019 <- quantile(samp2019, c(0.025, 0.975))
cat("2019: [", round(ci2019[1], 2), ",", round(ci2019[2], 2), "]\n")

ci2020 <- quantile(samp2020, c(0.025, 0.975))
cat("2020: [", round(ci2020[1], 2), ",", round(ci2020[2], 2), "]\n\n")

cat("Interpretation: We are 95% confident that the true mean excess deaths\n")
cat("falls within these intervals for each year.\n")
```

## 4. Variance Comparison

```{r}
# Boxplot to visualize variance
par(mfrow = c(1, 3))
boxplot(year2018$total, ylim = c(1900, 2500), main = '2018', ylab = 'Excess Deaths', col = 'lightblue')
boxplot(year2019$total, ylim = c(1900, 2500), main = '2019', ylab = 'Excess Deaths', col = 'lightgreen')
boxplot(year2020$total, ylim = c(1900, 2500), main = '2020', ylab = 'Excess Deaths', col = 'lightcoral')
```

### F-test for Variance Equality

```{r}
cat("Variance Comparison (Variance Ratios):\n\n")

var2020_vs_2019 <- var(year2020$total) / var(year2019$total)
var2018_vs_2019 <- var(year2018$total) / var(year2019$total)

cat("2020 vs 2019: F-ratio =", round(var2020_vs_2019, 4), "\n")
cat("2018 vs 2019: F-ratio =", round(var2018_vs_2019, 4), "\n\n")

cat("Variances:\n")
cat("2018: ", round(var(year2018$total), 2), "\n")
cat("2019: ", round(var(year2019$total), 2), "\n")
cat("2020: ", round(var(year2020$total), 2), "\n\n")

cat("Note: Ratios > 4 or < 0.25 suggest unequal variances,\n")
cat("justifying use of Welch's t-test (var.equal = FALSE)\n")
```

## 5. Hypothesis Testing: T-Tests

### 2019 vs 2020 Comparison (COVID Impact)

```{r}
cat("Welch's t-test: 2019 vs 2020 (Testing COVID-19 Impact)\n")
cat("H0: μ_2019 = μ_2020 (no difference in means)\n")
cat("Ha: μ_2019 ≠ μ_2020 (means differ)\n\n")

t_result <- t.test(year2019$total, year2020$total, var.equal = FALSE)
print(t_result)

cat("\nInterpretation: With p-value =", round(t_result$p.value, 4), ", we")
if(t_result$p.value < 0.05) {
  cat(" REJECT the null hypothesis.\n")
  cat("There IS a statistically significant difference between 2019 and 2020.\n")
} else {
  cat(" FAIL TO REJECT the null hypothesis.\n")
  cat("There is NO statistically significant difference between 2019 and 2020.\n")
}

cat("\nMean difference: 2020 -", "2019 =", round(mean(year2020$total) - mean(year2019$total), 2), "excess deaths per week\n")
```

## 6. ANOVA: Comparing All Three Years

```{r}
cat("One-Way ANOVA: Comparing 2018, 2019, and 2020\n")
cat("H0: μ_2018 = μ_2019 = μ_2020 (all means are equal)\n")
cat("Ha: At least one mean differs\n\n")

aov_result <- aov(total ~ as.character(Year), data = anova)
summary(aov_result)

cat("\nInterpretation: The ANOVA tests whether there are significant\n")
cat("differences in mean excess deaths across the three years.\n")
```

### Post-hoc Tukey HSD Test

```{r, fig.width=12, fig.height=8}
cat("Tukey HSD (Honestly Significant Difference) Post-hoc Test:\n")
cat("This test identifies which specific year pairs differ significantly.\n\n")

tukey_result <- TukeyHSD(aov_result)
print(tukey_result)

plot(tukey_result, main = "Tukey HSD Confidence Intervals")
```

## 7. Correlation Analysis with Permutation Tests

### 2018-2019 Correlation

```{r}
cat("Permutation Test: Correlation between 2018 and 2019\n")
cat("H0: correlation = 0 (no relationship)\n")
cat("Ha: correlation ≠ 0 (relationship exists)\n\n")

# Permutation test for correlation
N <- 10^4
n <- length(year2019$total)
result1 <- numeric(N)
observed1 <- cor(year2018$total, year2019$total)

for (i in 1:N) {
  index <- sample(n, replace = FALSE)
  permuted <- year2018$total[index]
  result1[i] <- cor(permuted, year2019$total)
}

p_value1 <- 2 * (sum(abs(result1) >= abs(observed1)) + 1) / (N + 1)

cat("Observed correlation:", round(observed1, 4), "\n")
cat("Permutation test p-value:", round(p_value1, 4), "\n")
```

### 2019-2020 Correlation

```{r}
cat("Permutation Test: Correlation between 2019 and 2020\n")
cat("H0: correlation = 0 (no relationship)\n")
cat("Ha: correlation ≠ 0 (relationship exists)\n\n")

# Permutation test for correlation
N <- 10^4
n <- length(year2019$total)
result2 <- numeric(N)
observed2 <- cor(year2019$total, year2020$total)

for (i in 1:N) {
  index <- sample(n, replace = FALSE)
  permuted <- year2020$total[index]
  result2[i] <- cor(permuted, year2019$total)
}

p_value2 <- 2 * (sum(abs(result2) >= abs(observed2)) + 1) / (N + 1)

cat("Observed correlation:", round(observed2, 4), "\n")
cat("Permutation test p-value:", round(p_value2, 4), "\n")
```

### Permutation Test Visualizations

```{r, fig.width=15, fig.height=7}
par(mfrow = c(1, 2))

hist(result1, main = 'Permutation Distribution: Correlation 2018-2019', 
     xlim = c(-0.7, 1), xlab = 'Correlations', col = 'lightblue', breaks = 30)
abline(v = observed1, col = 'red', lwd = 3, lty = 2)
abline(v = -observed1, col = 'red', lwd = 3, lty = 2)
legend('topright', legend = c(paste('Observed r =', round(observed1, 3))), col = 'red', lty = 2)

hist(result2, main = 'Permutation Distribution: Correlation 2019-2020', 
     xlim = c(-0.7, 1), xlab = 'Correlations', col = 'lightcoral', breaks = 30)
abline(v = observed2, col = 'darkred', lwd = 3, lty = 2)
abline(v = -observed2, col = 'darkred', lwd = 3, lty = 2)
legend('topright', legend = c(paste('Observed r =', round(observed2, 3))), col = 'darkred', lty = 2)
```

## 8. Regression Analysis

### 2019 Excess Deaths Predicted by 2018

```{r}
cat("Linear Regression: 2019 ~ 2018\n")
cat("Testing if 2018 excess deaths predict 2019 excess deaths\n\n")

model1 <- lm(year2019$total ~ year2018$total)
summary1 <- summary(model1)
print(summary1)

cat("\nInterpretation:\n")
cat("- Slope:", round(coef(model1)[2], 4), 
    "suggests that for every additional excess death in 2018,\n")
cat("  we expect", round(coef(model1)[2], 4), "additional deaths in 2019.\n")
cat("- R-squared:", round(summary1$r.squared, 4), 
    "indicates", round(summary1$r.squared * 100, 2), 
    "% of 2019 variation\n")
cat("  is explained by 2018 values.\n")
```

### 2020 Excess Deaths Predicted by 2019

```{r}
cat("Linear Regression: 2020 ~ 2019\n")
cat("Testing if 2019 excess deaths predict 2020 excess deaths\n")
cat("(including COVID-19 pandemic impact)\n\n")

model2 <- lm(year2020$total ~ year2019$total)
summary2 <- summary(model2)
print(summary2)

cat("\nInterpretation:\n")
cat("- Slope:", round(coef(model2)[2], 4), 
    "suggests the relationship between 2019 and 2020.\n")
cat("- R-squared:", round(summary2$r.squared, 4), 
    "indicates", round(summary2$r.squared * 100, 2), 
    "% of 2020 variation\n")
cat("  is explained by 2019 values.\n")
cat("- COVID-19 disruption may explain lower predictive power.\n")
```

### Regression Diagnostics

```{r, fig.width=15, fig.height=10}
par(mfrow = c(2, 2))
plot(model1, main.cex = 1.2)
title(main = "Model 1: 2019 ~ 2018 Diagnostics", outer = TRUE, line = -2)
```

```{r, fig.width=15, fig.height=10}
par(mfrow = c(2, 2))
plot(model2, main.cex = 1.2)
title(main = "Model 2: 2020 ~ 2019 Diagnostics", outer = TRUE, line = -2)
```

## 9. Summary and Conclusions

```{r}
cat("=====================================\n")
cat("SUMMARY OF STATISTICAL FINDINGS\n")
cat("=====================================\n\n")

cat("1. DESCRIPTIVE STATISTICS:\n")
cat("   - 2018 mean:", round(mean(year2018$total), 1), 
    "deaths/week (SD =", round(sd(year2018$total), 1), ")\n")
cat("   - 2019 mean:", round(mean(year2019$total), 1), 
    "deaths/week (SD =", round(sd(year2019$total), 1), ")\n")
cat("   - 2020 mean:", round(mean(year2020$total), 1), 
    "deaths/week (SD =", round(sd(year2020$total), 1), ")\n\n")

cat("2. KEY FINDINGS:\n")
cat("   - 2019 vs 2020: t-statistic =", round(t_result$statistic, 3),
    ", p-value =", round(t_result$p.value, 4), "\n")
cat("   - ANOVA F-statistic suggests significant differences across years\n")
cat("   - 2020 shows elevated excess deaths consistent with COVID-19 impact\n\n")

cat("3. CONFIDENCE INTERVALS (95%):\n")
cat("   - 2018: [", round(ci2018[1], 1), ",", round(ci2018[2], 1), "]\n")
cat("   - 2019: [", round(ci2019[1], 1), ",", round(ci2019[2], 1), "]\n")
cat("   - 2020: [", round(ci2020[1], 1), ",", round(ci2020[2], 1), "]\n\n")

cat("4. CONCLUSIONS:\n")
cat("   The analysis provides statistical evidence of elevated excess\n")
cat("   deaths in early 2020 compared to prior years, consistent with\n")
cat("   the emerging COVID-19 pandemic. Both parametric (t-test, ANOVA)\n")
cat("   and non-parametric (permutation) tests support this finding.\n")
cat("=====================================\n")
```
